Stateless Model Checking with Data-Race Preemption Points

### 2015-11-17 pldi-16/design.tex
Move a lot of stuff from i to d

Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
Iterative Deepening likewise makes progressively deeper searches of the state space until the CPU budget is exhausted.
In our context, the ``depth'' is the number of PPs used.
We rely on state-space estimation \cite{estimation}
to understand which state spaces are likely to complete on time,
in advance of actually testing each interleaving within.
The overall purpose is to make decisions automatically about when to defer exploration of a state space,
so an inexpert user must provide only their total CPU budget as a test parameter,
and to enable completing appropriately-sized state spaces within that budget.

Note that Iterative Deepening is a {\em wrapper} algorithm around stateless model checking.
A model checking tool is still used to test each state space, and other reduction techniques are still applicable.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new ones reactively based on any runtime analysis.
In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.

@@

\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
\newcommand\GetETA[1]{ETA(#1)}
\newcommand\GetPPSet[1]{PPSet(#1)}
With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
and a set of {\em pending} jobs (untested ones with unknown ETAs).
When a MC instance reports an ETA too high,
we compare other pending jobs to find another one more likely to complete in time.
%
Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening\footnote{
Though its worst-case performance is $O(mn)$ in the
%number of pending and suspended jobs,
sizes of $\mathcal{P}$ and $\mathcal{S}$,
in practice the non-constant portion beyond line 4 runs very infrequently
and is negligible compared to the exponentially-sized state spaces.}.
Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
and $j_1$ is suspended,
then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
%reveals that it might finish in time after all.
Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.

@@

\subsection{Classifying data races as buggy or benign}
\label{sec:classifying}

Iterative Deepening must be seeded with some initial state spaces,
which can be any number of subsets of the statically-available PPs that SSS-MC would use.
While exploring those state spaces, runtime data-race detection may find data-race candidates that we wish to investigate further.
Because data races indicate access pairs that can interleave at instruction granularity,
it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.

@@

@@

>>>>

Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
Iterative Deepening
%likewise
makes progressively deeper searches of the state space until the CPU budget is exhausted.
In this context, the depth is the number of PPs used.
Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs
, which we refer to as {\em jobs}.
It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.

We rely on state-space estimation \cite{estimation}
to understand which jobs are likely to complete on time,
before actually testing each interleaving within.
The overall purpose is to decide automatically when to defer testing a state space,
so an inexpert user can provide only their total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.

Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
A MC tool is still used to test each state space, and other reduction techniques are still applicable.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new ones reactively based on any runtime analysis.
We focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Initial PP configuration}

Iterative Deepening must be seeded with some initial state spaces,
which can be any number of subsets of the statically-available PPs that SSS-MC would use.
For testing user-space code, we begin with the four subsets of locking API PPs that we showed in Figure~\ref{fig:id}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also introduce PPs on any other sync primitives implemented with internal locks,
such as condvars or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
%so the {\tt yield} PP is always implicitly enabled.

For kernel-level testing, we consider interrupt-disabling analogous to locking,
so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
%(to appropriate the names of the x86 instructions)
During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
Some kernels disable preemption without disabling interrupts,
which can be modelled the same way using manual annotations. %of that API.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
\quicksand~is configured to begin with
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, we don't test every intermediate subset such as $\{lock,sti\}$,
which could potentially be improved in future work (\sect{\ref{sec:future}}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
\newcommand\GetETA[1]{ETA(#1)}
\newcommand\GetPPSet[1]{PPSet(#1)}
With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
and a set of {\em pending} jobs (untested ones with unknown ETAs).
When a MC instance reports an ETA too high,
we compare other pending jobs to find another one more likely to complete in time.
%
Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
%\footnote{
%Though its worst-case performance is $O(mn)$ in the
%%number of pending and suspended jobs,
%sizes of $\mathcal{P}$ and $\mathcal{S}$,
%in practice the non-constant portion beyond line 4 runs very infrequently
%and is negligible compared to the exponentially-sized state spaces.}.
Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
and $j_1$ is suspended,
then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
%reveals that it might finish in time after all.
Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Data-race preemption points}
\label{sec:classifying}

During stateless MC, runtime data-race detection may find data-race candidates that we wish to investigate further.
Because data races indicate access pairs that can interleave at instruction granularity,
it is logical to re-execute the test and issue preemptions during those instructions to test alternate thread interleavings~\cite{racefuzzer,portend}.

@@

Furthermore, we do not always increase the number of PPs when we find a new data race.
%When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
For each instruction involved in a data race, \quicksand~adds two new jobs:
a ``small'' job to preempt on that instruction only,
and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
%
Hence,
%together with the logic in \sect{\ref{sec:classifying}},
each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
%
The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
while the big job risks not completing in time,
the small job risks missing the data race entirely if the original PPs were required to expose it.
% TODO CAMREADY: Put numbers here.
In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
which motivates the need for Iterative Deepening to prioritize the jobs at runtime.

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{dr-jobs-v2.pdf}
	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
	\label{fig:new-dr-jobs}
\end{figure}

@@

%When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
%Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
%(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
%Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
%Instead, \landslide~qualifies the PC with
%(a) the current thread ID and
%% FIXME: We don't actually do this.
%(b) the most recent {\tt call} instruction.
%% (a crude approximation of a stack trace)
%% which are much cheaper, as we carry them around all the time already
%Note that we do {\em not} qualify data races by the shared memory address,
%which can change based on different interleavings of previous code
%(for example, depending on the result of {\tt malloc()}).
%% especially when malloc is involved.
%%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Heuristics}
% List of all heuristix:
% HOMESTRETCH - last 60sec of test, don't suspend
% ETA_THRESH - "to let its ETA stabilize"
% eta factor
% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
% priority change between suspected and confirmed dr
Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
to express how optimistic we are about the estimate's accuracy.
We use 2 as this scaling factor based on the results in \cite{estimation}.
%though we allow changing it via the command line.
We also include a heuristic to
%ignore ETAs entirely
never suspend jobs before they pass a certain threshold of interleavings tested,
for which we choose 32,
so that their ETAs have some time to stabilize.

We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
Jobs with both-order data-race PPs are prioritized higher,
because single-order candidates are more likely to be false-positives
(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
%so we prioritize jobs with both-order data-race PPs.
For single-order races, we do not add a PP for the later access at all;
should it be needed, preempting on the first access will suffice to upgrade the race to both-order.

### 2015-11-17 pldi-16/implementation.tex
Fix pictures
Move a lot of stuff from i to d

We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
\landslide~implements DPOR \cite{dpor},
%\cite{dpor},
state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
% TODO: \cite{fairstatelessmc} instead or in addition?
It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
% this line can be cut if space is needed
It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.

{\bf Restricting PPs with stack trace predicates.}
Most model checkers (MCs) preempt indiscriminately on any sync API call, regardless of the call-site.
However, when testing a particular module in a large codebase,
the user is likely uninterested in PPs arising from other modules.
\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
%The {\tt without\_function} directive works similarly, but as a blacklist.
The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
Multiple invocations can be used; later ones take precedence.
%\cite{landslide} provides further detail on this feature.

@@

\quicksand~is an independent program that wraps the execution of several \landslide~instances.
The implementation is roughly 3000 lines of C.
It uses a thread pool to schedule the available state spaces,
sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.

{\bf Initial PPs.}
For testing user-space code, we seed the exploration with the four subsets of ``hard-coded'' locking API PPs that we showed in Figure~\ref{fig:id}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also introduce preemptions on any other sync primitives implemented with mutexes,
such as condition variables, semaphores, and r/w locks.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
%so the {\tt yield} PP is always implicitly enabled.

For kernel-level testing, we consider interrupt-disabling to be analogous to locking,
so we also preempt just before a disable-interrupt opcode ({\tt cli}) and just after interrupts are re-enabled ({\tt sti})\footnote{
%(to appropriate the names of the x86 instructions)
During data-race detection, {\tt cli}/{\tt sti} are treated as a single global lock.
%as {\tt cli}'d memory accesses can still race with others that have interrupts on.
Some kernels disable preemption without disabling interrupts,
which can be modelled the same way using manual annotations. %of that API.
This also assumes uni-processor scheduling; for SMP kernels, replace {\tt cli}/{\tt sti} with spinlocks.}.
\quicksand~is configured to begin with the subsets
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
$\{yield,cli\}$,
$\{yield,sti\}$,
and $\{yield,lock,$ $unlock,cli,sti\}$.
As a heuristic, we don't bother with every intermediate subset such as $\{lock,sti\}$,
which could potentially be improved in future work (\sect{\ref{sec:future}}).

{\bf Communication protocol.}
The interface to \landslide~, which any similar MC could implement, has two parts.
First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
% can lose this line due to space
among other options such as mutex-testing mode,
passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
whenever a new data-race candidate is found, and whenever a bug is found.
\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
We implement suspending jobs simply by making \landslide~wait on a message-passing call.
Should \quicksand~later resume a suspended job, we send a message to continue,
causing the \landslide~instance to resume exploring where it left off;
otherwise, we send a message only after time runs out, causing it to exit.

{\bf Heuristics.}
% List of all heuristix:
% HOMESTRETCH - last 60sec of test, don't suspend
% ETA_THRESH - "to let its ETA stabilize"
% eta factor
% shold_reproduce -- small dr jobs are not allowed to add further instances of themself (why not? don't remember)
% priority change between suspected and confirmed dr
Algorithm~\ref{alg:shouldworkblock} allows heuristically scaling a job's ETA when comparing to the time budget,
to express how optimistic we are about the estimate's accuracy.
We use 2 as this scaling factor based on the results in \cite{estimation},
though we allow changing it via the command line.
We also include a heuristic to
%ignore ETAs entirely
never suspend jobs before they pass a certain threshold of interleavings tested,
for which we choose 32,
so that their ETAs have some time to stabilize.

We classify data-race candidates as {\em single-order} or {\em both-order} \cite{portend}
based on whether the MC observed the racing instructions ordered one or both ways in the original state space,
Jobs with both-order data-race PPs are prioritized higher,
because single-order candidates are more likely to be false-positives
(though before preempting during the access itself, we cannot say for sure, hence the heuristic).
%so we prioritize jobs with both-order data-race PPs.
For single-order races, we do not add a PP for the later access at all;
should it be needed, preempting on the first access will suffice to upgrade the race to both-order.

% If there's room, mention the cant_swap mechanism for killing the top half of deferred jobs.

\subsection{Data-race preemption points}

When \landslide~detects a data race, it reports each of the two memory accesses involved in the race.
Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
(For example, many parts of a codebase might call {\tt list\_insert()}, but only one callsite does so without adequate locking.)
Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do for each shared memory access.
Instead, \landslide~qualifies the PC with
(a) the current thread ID and
% FIXME: We don't actually do this.
(b) the most recent {\tt call} instruction.
% (a crude approximation of a stack trace)
% which are much cheaper, as we carry them around all the time already
Note that we do {\em not} qualify data races by the shared memory address,
which can change based on different interleavings of previous code
(for example, depending on the result of {\tt malloc()}).
% especially when malloc is involved.
%Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{dr-jobs.pdf}
	\caption{\quicksand~manages the exploration of multiple state spaces, communicating with each MC instance to receive ETAs, data race candidates, and bug reports.
		When an access pair is reported as a data race, we generate a new PP for each access and add new jobs corresponding to different combinations of those with the existing PPs.}
	\label{fig:new-dr-jobs}
\end{figure}

When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
a ``small'' job to preempt on the racing instruction only,
and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
%
Hence, together with the logic in \sect{\ref{sec:classifying}}, each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
%
The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance:
while the big job risks not completing in time,
the small job risks missing the data race entirely if the original PPs were required to expose it.
In practice, we observed some bugs found quickly by these small jobs, and other bugs missed by the small jobs found eventually by the big jobs,
which motivates the need for Iterative Deepening to prioritize the jobs at runtime.
% TODO: Put numbers here.

>>>>

We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
\landslide~implements DPOR \cite{dpor},
%\cite{dpor},
state space estimation \cite{estimation}, and a hybrid lockset/happens-before data-race analysis \cite{hybriddatarace}.
It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}.
% this line can be cut if space is needed
%It can test both user-level and kernel-level code, although is limited to timer-driven nondeterminism.
% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.

{\bf Restricting PPs with stack trace predicates.}
Most MC tools preempt indiscriminately on any sync API call, regardless of the call-site.
However, when testing a particular module in a large codebase,
the user is likely uninterested in PPs arising from other modules.
\landslide~provides the {\tt within\_function} configuration command for a user to identify which call-sites matter most.
Before inserting a PP, \landslide~requires at least one argument to {\tt within\_function} to appear in the current thread's stack trace.
%The {\tt without\_function} directive works similarly, but as a blacklist.
The {\tt without\_function} directive is the dual of {\tt within\_function}, indicating a blacklist.
Multiple invocations can be used; later ones take precedence.
%\cite{landslide} provides further detail on this feature.

@@

\quicksand~is an independent program that wraps the execution of several \landslide~instances.
The implementation is roughly 3000 lines of C.
%It uses a thread pool to schedule the available state spaces,
%sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
%Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
The interface to \landslide~ has two parts. %, which any similar MC could implement, has two parts.
First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
% can lose this line due to space
among other options such as mutex-testing mode,
passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing. %on a FIFO pipe.
\landslide~messages \quicksand~after testing each new interleaving to report updated progress and ETA,
whenever a new data-race candidate is found, and whenever a bug is found.
\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.
We implement suspending jobs simply by making \landslide~wait on a message-passing call.
Should \quicksand~later resume a suspended job, we send a message to continue,
resuming the \landslide~instance where it left off;
otherwise, we resume it only after time runs out, causing it to exit.
